Robust DPG stability II

### 2012-04-02 docs/NewBCRobustness_paper/paper.tex
made more changes (proofreading) to paper
git-svn-id: https://svn.ices.utexas.edu/repos/camellia@423 36274a68-8941-4c93-80ef-affaed5139da

Standard Bubnov-Galerkin methods tend to perform poorly for the class of partial-differential equations known as singularly perturbed problems, where a given parameter $\epsilon$ may approach either $0$ or $\infty$ in the context of physical problems.  This poor performance is captured by the error bound
\[
\|u-u_h\|_E \leq C(\epsilon) \inf_{w_h}\|u-w_h\|_U
\]
The growth of $C$ with $\epsilon$ is referred to as a loss of robustness, where the bound on the error in the finite element solution by the best approximation error increases as $\epsilon\rightarrow 0$.  Intuitively, as our singular perturbation parameter changes, our finite element error is bounded more and more loosely to the best approximation error, allowing for degradation of the solution

@@

Traditionally, instability/loss of robustness has been dealt with using residual-based stabilization.  Given some variational form $b(u,v) - l(v)=0$, the problem is modified by adding to the bilinear form a factor of the residual $r(v) = b(u,v) - l(v)=0$, scaled by a stabilization constant $\tau$.  The most well-known example of this is the streamline-upwind Petrov-Galerkin method, which is a stabilized method for solving the convection-diffusion equation using piecewise linear continuous finite elements \cite{SUPG}.

@@

The DPG (discontinuous Petrov-Galerkin) method is the combination of the concept of computable optimal test functions with the so-called ``ultra-weak formulation", where the regularity requirement on each solution variable is relaxed through integration by parts.  Moreover, to maintain conformity while seeking an $L^2$ setting on interior ``field" variables, boundary terms are identified as additional new unknowns.  The result is a hybridized DG method with locally supported optimal test functions.  For a given operator equation $Au = f$, the ultra-weak formulation is 
\[
b\left(\left(\widehat{u},u\right),v\right) - l(v) = \langle \widehat{u}, [v] \rangle - (u,A^*v) - (f,v) = 0
\]
where $A^*$ is the formal adjoint resulting from integration by parts, and $[v]$ is the jump of $v$ across element boundaries.

@@

The duality between test norm and variational form holds for any particular variational form, though it is often difficult to determine analytically the form of $\|v\|_{V,U}$.  However, for the DPG ultra-weak formulation, there is an important test norm that results from a specific induced norm on $U$.  
Define the norm on $U$
\[
\|u\| = C_1\|u\|_{L^2} + C_2\|\widehat{u}\|
\]
where the norm on $\widehat{u}$ is taken to be the minimum energy extension norm
\[
\|\widehat{u}\| = \inf_{w\in V, \left.w\right|_{\Gamma_h}=\widehat{u}} \|w\|_V
\]
Then, the induced test norm is 
\begin{align*}
\|v\|_{\rm opt} &= \sup_{u,\widehat{u}} \frac{\langle \widehat{u}, [v] \rangle - (u,A^*v)}{C_1\|u\|_{L^2} + C_2\|\widehat{u}\|} \\
&\leq \frac{1}{C_1}\sup_{u} \frac{(u,A^*v)}{\|u\|_{L^2}} + \frac{1}{C_2}\sup_{\widehat{u}} \frac{\langle \widehat{u}, [v] \rangle}{\|\widehat{u}\|}\\
&= \frac{1}{C_1}\|A^*v\|_{L^2} + \frac{1}{C_2}\sup_{\widehat{u}} \frac{\langle \widehat{u}, [v] \rangle}{\|\widehat{u}\|}
\end{align*}
where we have used the Cauchy-Schwarz inequality, and eliminated the supremum by noting the inequality is tight.  We define $\|u\|_{\rm opt}$ as such a norm with $C_1=C_2 = 1$,
\[
\|u\|_{\rm opt} = \|u\|_{L^2} + \|\widehat{u}\|.
\]
This is, in some sense, the ideal or ``optimal" test norm; given any particular norm $\|u\|_U$, DPG under the optimal test norm will return the best finite element approximation in both field and flux variables, independently of any parameters

The norm $\|v\|_{\rm opt}$ is non-localizable, however - the above ``optimal" test norm is induced by an inner product; however, this inner product is a global one due to the presence of the jump terms.  Solving for optimal test functions under this optimal norm through inversion of this inner product  will result in expensive global problems, making this an impractical norm to work with.

@@

It's worth addressing the effect of boundary conditions on stability.  Specifically, a test norm that provides stability for one set of boundary conditions may perform poorly for another set of boundary conditions.  The bilinear form for the case of Dirichlet boundary conditions is 
\[
b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right) = \left(u,\div \tau - \beta \cdot \grad v\right) + \left(\sigma, \epsilon^{-1} \tau + \grad v\right) + \langle \widehat{u}, \left[\tau \cdot n\right] \rangle_{\Gamma_h^0} + \langle \widehat{f}_n, \left[v\right] \rangle_{\Gamma_h}
\]
Notice that the final bilinear form is different; hence, the adjoint problems associated with Section~\ref{sec:strategy3} will now carry different boundary conditions as well.  The stability properties proven here 
%We note that 
%\begin{align*}
%b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right) &= \left(u,\div \tau - \beta \cdot \grad v\right) + \left(\sigma, \epsilon^{-1} \tau + \grad v\right) - \langle \left[\tau_n\right], \widehat{u} \rangle_{\Gamma_h^0} + \langle \widehat{f}_n, \left[v\right] \rangle_{\Gamma_h} 
%&= \| u \|^2_{L^2}
%\end{align*}
%for the specific continuous test functions $\tau$ and $v$ satisfying
%\begin{align*}
%\div \tau - \beta \cdot \grad v &= u \\
%\frac{1}{\epsilon}\tau + \grad v &= 0
%\end{align*}
%with boundary condition $\left.v\right|_\Gamma = 0$. Choosing a continuous conforming test function $v$ removes the contribution of the term $\langle \widehat{f}_n, \left[v\right] \rangle$ on the mesh skeleton, while the boundary conditions remove the contribution of boundary terms from the bilinear form.  Then, for this specific choice of $\left(\tau, v\right)$, 
%\begin{align*}
%\|u\|_{L^2}^2 &= b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right) = \frac{b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right)}{\|\left(\tau, v\right)\|_V}\|\left(\tau, v\right)\|_V \leq \|u\|_E \|\left(\tau, v\right)\|_V
%\end{align*}
%At minimum, we want $\|u\|_{L^2} \lesssim \|u\|_E$, which requires 
%\[
%\|\left(\tau, v\right)\|_V \lesssim \| u \|_{L^2}
%\]
%for any choice of $u \in L^2$ (dividing through by $\|u\|_{L^2}$ gives the desired result).  This can be considered a ``sanity check"; at the minimum, we at least want control over the behavior of the single field variable $u$.  However, we are unable to prove this in the case of Dirichlet boundary conditions. 
The robust bounds given in Section~\ref{sec:main_bounds} hold in arbitrary dimension; however, we can show that for the case of Dirichlet boundary conditions, the same results will not hold, even in 1D.

@@

Consider now the 1D analogue of the estimate given by Lemma~\ref{lemma_stream}, which is necessary for robust control over the error in $\|u\|_{L^2}$.  In 1D, $\|\beta\cdot\grad v\| \lesssim \|g\|$ reduces to the inequality 
\[
\|v'\|\lesssim \|g\|, \quad g\in L^2{\Omega_h}
\]
The adjoint problem corresponding to Lemma~\ref{lemma_stream} in section~\ref{sec:strategy3} is likewise reduced in 1D to the scalar equation
\begin{align}
\epsilon v'' + v' &= -g \label{adjoint_1D}
\end{align}
with $v\in H^1_0\left((0,1)\right)$.  Multiplying this equation by $v'$ and integrating over $\Omega_h$ gives
\[
\int_0^1 \epsilon v''v' + \|v'\|_{L^2}^2 = -\int_0^1 g v'
\]
Integrating by parts and Young's inequality give
\begin{align*}
-\left.\frac{\epsilon}{2} v'^2\right|_0^1 + \|v'\|_{L^2}^2 &\leq \frac{1}{2}\|g\| + \frac{1}{2}\|v'\|
\end{align*}
implying that
\begin{align*}
\frac{\epsilon}{2} v'(0)^2 + \|v'\|_{L^2}^2 & \lesssim \|g\| 
\end{align*}
Let us restrict ourselves to the cases where $v$ is sufficiently smooth for $v'(0)$ to be well defined.  Taking $g=1$, we can solve \eqref{adjoint_1D} exactly to see that $\epsilon v'(0)^2 = O(\epsilon^{-1})$, indicating a lack of robustness\footnote{Unlike the case of Dirichlet boundary conditions, the inflow condition on $ \widehat{f} = u(0)-\epsilon u'(0)$ induces an adjoint boundary condition $\tau(0)=0$, or equivalently $v'(0) = 0$, removing the non-robust term from the estimate.}.

>>>>

Standard Bubnov-Galerkin methods tend to perform poorly for the class of partial-differential equations known as singularly perturbed problems, where a given parameter $\epsilon$ may approach either $0$ or $\infty$ in the context of physical problems.  This poor performance is captured by the error bound
\[
\|u-u_h\|_E \leq C(\epsilon) \inf_{w_h}\|u-w_h\|_U
\]
The growth of $C$ with $\epsilon$ is referred to as a loss of robustness, where the bound on the error in the finite element solution by the best approximation error increases as $\epsilon\rightarrow 0$.  Intuitively, as our singular perturbation parameter changes, our finite element error is bounded more and more loosely by the best approximation error, allowing for degradation of the solution for very small $\epsilon$.

@@

Traditionally, instability/loss of robustness has been dealt with using residual-based stabilization.  Given some variational form, the problem is modified by adding to the bilinear form a factor of the residual, scaled by a stabilization constant $\tau$.  The most well-known example of this is the streamline-upwind Petrov-Galerkin method, which is a stabilized method for solving the convection-diffusion equation using piecewise linear continuous finite elements \cite{SUPG}.

@@

The DPG (discontinuous Petrov-Galerkin) method is the pairing of the concept of computable optimal test functions with the so-called ``ultra-weak formulation", where the regularity requirement on each solution variable is relaxed through integration by parts.  Moreover, to maintain conformity while seeking an $L^2$ setting on interior ``field" variables, boundary terms are identified as additional new unknowns.  The result is a hybridized DG method with locally supported optimal test functions.  For a given operator equation $Au = f$, the ultra-weak formulation is 
\[
b\left(\left(\widehat{u},u\right),v\right) - l(v) = \langle \widehat{u}, [v] \rangle - (u,A^*v) - (f,v) = 0
\]
where $A^*$ is the formal adjoint resulting from integration by parts, and $[v]$ is the jump of $v$ across element boundaries.

@@

The duality between test norm and variational form holds for any particular variational form, though it is often difficult to determine analytically the form of $\|v\|_{V,U}$.  However, for the DPG ultra-weak formulation, there is an important test norm that results from a specific induced norm on $U$.  
Define the norm on $U$
\[
\|u\| = C_1\|u\|_{L^2} + C_2\|\widehat{u}\|
\]
where the norm on $\widehat{u}$ is taken to be the minimum energy extension norm
\[
\|\widehat{u}\| = \inf_{w\in V, \left.w\right|_{\Gamma_h}=\widehat{u}} \|w\|_V
\]
Then, the induced test norm is 
\begin{align*}
\|v\| &= \sup_{u,\widehat{u}} \frac{\langle \widehat{u}, [v] \rangle - (u,A^*v)}{C_1\|u\|_{L^2} + C_2\|\widehat{u}\|} \\
&\leq \frac{1}{C_1}\sup_{u} \frac{(u,A^*v)}{\|u\|_{L^2}} + \frac{1}{C_2}\sup_{\widehat{u}} \frac{\langle \widehat{u}, [v] \rangle}{\|\widehat{u}\|}\\
&= \frac{1}{C_1}\|A^*v\|_{L^2} + \frac{1}{C_2}\sup_{\widehat{u}} \frac{\langle \widehat{u}, [v] \rangle}{\|\widehat{u}\|}
\end{align*}
where we have used the Cauchy-Schwarz inequality, and eliminated the supremum by noting the inequality is tight.  We define $\|u\|_{\rm opt}$ as such a norm with $C_1=C_2 = 1$,
\[
\|u\|_{\rm opt} = \|u\|_{L^2} + \|\widehat{u}\|.
\]
which induces the so-called ``optimal test norm"
\[
\|v\|_{\rm opt} = \|A^*v\|_{L^2} + \left(\sup_{\widehat{u}} \frac{\langle \widehat{u}, [v] \rangle}{\|\widehat{u}\|}\right)
\]
This is, in some sense, the canonical or ideal test norm; given any well-posed variational problem, DPG under the optimal test norm will return the best finite element approximation in both field and flux variables.

The norm $\|v\|_{\rm opt}$ is non-localizable, however - the optimal test norm is induced by a global inner product due to the presence of the jump terms.  Solving for optimal test functions under this optimal norm (through inversion of Riesz map correspdongin to this inner product) will result in expensive global problems, making this an impractical norm to work with.

@@

It's worth addressing the effect of boundary conditions on stability.  Specifically, a test norm that provides stability for one set of boundary conditions may perform poorly for another set.  Take, for example, the test norm defined in Section~\ref{sec:main_bounds} and the convection-diffusion problem with Dirichlet boundary conditions.

The bilinear form for the case of Dirichlet boundary conditions is 
\[
b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right) = \left(u,\div \tau - \beta \cdot \grad v\right) + \left(\sigma, \epsilon^{-1} \tau + \grad v\right) + \langle \widehat{u}, \left[\tau \cdot n\right] \rangle_{\Gamma_h^0} + \langle \widehat{f}_n, \left[v\right] \rangle_{\Gamma_h}
\]
Notice that the final bilinear form is different; hence, the adjoint problems associated with Section~\ref{sec:strategy3} will now carry different boundary conditions as well.  The stability properties proven here 
%We note that 
%\begin{align*}
%b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right) &= \left(u,\div \tau - \beta \cdot \grad v\right) + \left(\sigma, \epsilon^{-1} \tau + \grad v\right) - \langle \left[\tau_n\right], \widehat{u} \rangle_{\Gamma_h^0} + \langle \widehat{f}_n, \left[v\right] \rangle_{\Gamma_h} 
%&= \| u \|^2_{L^2}
%\end{align*}
%for the specific continuous test functions $\tau$ and $v$ satisfying
%\begin{align*}
%\div \tau - \beta \cdot \grad v &= u \\
%\frac{1}{\epsilon}\tau + \grad v &= 0
%\end{align*}
%with boundary condition $\left.v\right|_\Gamma = 0$. Choosing a continuous conforming test function $v$ removes the contribution of the term $\langle \widehat{f}_n, \left[v\right] \rangle$ on the mesh skeleton, while the boundary conditions remove the contribution of boundary terms from the bilinear form.  Then, for this specific choice of $\left(\tau, v\right)$, 
%\begin{align*}
%\|u\|_{L^2}^2 &= b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right) = \frac{b\left(\left(u,\sigma, \widehat{u}, \widehat{\sigma}_n\right), \left(\tau, v\right)\right)}{\|\left(\tau, v\right)\|_V}\|\left(\tau, v\right)\|_V \leq \|u\|_E \|\left(\tau, v\right)\|_V
%\end{align*}
%At minimum, we want $\|u\|_{L^2} \lesssim \|u\|_E$, which requires 
%\[
%\|\left(\tau, v\right)\|_V \lesssim \| u \|_{L^2}
%\]
%for any choice of $u \in L^2$ (dividing through by $\|u\|_{L^2}$ gives the desired result).  This can be considered a ``sanity check"; at the minimum, we at least want control over the behavior of the single field variable $u$.  However, we are unable to prove this in the case of Dirichlet boundary conditions. 
The robust bounds given in Section~\ref{sec:main_bounds} hold in arbitrary dimension; however, we can show that for the case of Dirichlet boundary conditions, the same results will not hold, even in 1D.

@@

Consider now the 1D analogue of the estimate given by Lemma~\ref{lemma_stream}, which is necessary for robust control over the error in $\|u\|_{L^2}$.  In 1D, $\|\beta\cdot\grad v\| \lesssim \|g\|$ reduces to the inequality 
\[
\|v'\|\lesssim \|g\|, \quad g\in L^2{\Omega_h}
\]
The adjoint problem corresponding to Lemma~\ref{lemma_stream} in section~\ref{sec:strategy3} is likewise reduced in 1D to the scalar equation
\begin{align}
\epsilon v'' + v' &= -g \label{adjoint_1D}
\end{align}
with $v\in H^1_0\left((0,1)\right)$.  Multiplying this equation by $v'$ and integrating over $\Omega_h$ gives
\[
\int_0^1 \epsilon v''v' + \|v'\|_{L^2}^2 = -\int_0^1 g v'
\]
Integrating by parts and Young's inequality give
\begin{align*}
-\left.\frac{\epsilon}{2} v'^2\right|_0^1 + \|v'\|_{L^2}^2 &\leq \frac{1}{2}\|g\| + \frac{1}{2}\|v'\|
\end{align*}
implying that
\begin{align*}
\frac{\epsilon}{2} v'(0)^2 + \|v'\|_{L^2}^2 & \lesssim \|g\| 
\end{align*}
Let us restrict ourselves to the cases where $v$ is sufficiently smooth for $v'(0)$ to be well defined.  Taking $g=1$ (corresponding to a piecewise constant approximation) we can solve \eqref{adjoint_1D} exactly to see that 
\[
v = \frac{e^{-\frac{x}{\epsilon}}}{e^{\frac{1}{\epsilon}}-1}\left(e^{\frac{1}{\epsilon}}\left(e^{\frac{x}{\epsilon}}-1\right) + \left(e^{\frac{1}{\epsilon}}-1\right)e^{\frac{x}{\epsilon}}x\right)
\]
Consequently, $\frac{\epsilon}{2} v'(0)^2 \approx \epsilon^{-1}$, indicating already a lack of robustness in the bound of $\|v'\|$ by $\|g\|$ for constant data\footnote{Unlike the case of Dirichlet boundary conditions, the inflow condition on $ \widehat{f} = u(0)-\epsilon u'(0)$ induces an adjoint boundary condition $\tau(0)=0$, or equivalently $v'(0) = 0$, removing the non-robust term from the estimate.}. The robust bound from below of $g$ constant corresponds to the robust control of the approximation of field variable $u$ by piecewise constant trial functions. More detailed 1D error bounds for Dirichlet boundary conditions are provided in \cite{DPG3}, and also indicate a lack of robustness when using the standard broken Sobolev norm as the test norm on $\tau$ and $v$.
